---
title: "Getting Started with sigRecon"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sigRecon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The signature recontextualization problem

Given a gene signature $X$ describing a genetic or chemical perturbation in one
biological context (a cell line, a tissue), what is the corresponding gene
signature of that *same* perturbation in a *different* context? Any method
that takes a ranked list of genes from a source context and outputs a ranked
list of genes for a target context is performing **signature
recontextualization**.

`sigrecon` provides three baseline recontextualization methods --
projection-based scoring (`projectCor()`), network propagation
(`netProp()`), and a DESeq2-based mean method (`recontextualize(method =
"mean")`) -- plus benchmarking tools (`sig_eval_table()`) to evaluate how
well a recontextualized signature recovers the true signature in the target
context.

This vignette walks through all three methods end to end on a small, bundled
real-data example, requiring no downloads or external setup.

## Installation

```r
BiocManager::install("montilab/sigrecon", dependencies = TRUE)
```

```{r setup, message=FALSE}
library(sigrecon)
library(SummarizedExperiment)
```

## Demo Data

`sigrecon` bundles small, real-data recontextualization demos for each of the
datasets used in its benchmarking study (SciPlex, Tahoe, DrugMatrix,
Perturb-seq) -- see `?demo_datasets` for all four. This vignette uses the
SciPlex demo data:

- **`demo_sciplex_sigs`** -- SciPlex K562 signatures (top differentially
  expressed genes per drug), treated here as **source** signatures, as if
  defined in a different biological context than the one we're
  recontextualizing into.
- **`demo_sciplex_se`** -- SciPlex A549 expression data, representing
  the **target** context we want to recontextualize the source signatures
  into.
- **`demo_sciplex_true_sigs`** -- SciPlex A549 signatures for the same
  drugs, used as **ground truth** to check how well recontextualization
  worked.

```{r explore-demo}
dim(demo_sciplex_se)
length(demo_sciplex_sigs)
demo_sciplex_sigs[["Panobinostat (LBH589)"]][1:5]
```

Note `demo_sciplex_se` contains a *disjoint* set of drug profiles from those in
`demo_sciplex_sigs` and `demo_sciplex_true_sigs` -- so nothing
about the specific perturbations you're trying to recontextualize leaks into
the expression data used to predict new signatures.

## Method 1: Projection-based

`projectCor()` predicts a signature in a new context by scoring each
sample against the source signature (via GSVA or eigengene scoring) and then
ranking genes by their correlation with that per-sample score. The
highest-correlated genes become the recontextualized signature.

```{r projectcor}
recon_projectcor <- projectCor(demo_sciplex_se, demo_sciplex_sigs, score = "gsva")

# One recontextualized signature per source signature
length(recon_projectcor)
recon_projectcor[["Panobinostat (LBH589)"]][1:5]
```

## Method 2: Network Propagation

Network propagation works differently: `netProp()` first learns a gene
co-expression network from the target-context expression data, then
propagates the source signature across that network (via random walk with
restart) to find the genes most "reachable" from the seed genes -- these
become the recontextualized signature.

```{r netprop, message=FALSE}
recon_netprop <- netProp(demo_sciplex_se, seeds = demo_sciplex_sigs, sig = "rwr")
recon_netprop[["Panobinostat (LBH589)"]][1:5]
```

`netProp()` is a one-call wrapper around two lower-level functions,
`wgcna.adj()` (builds the network) and `network_sig()` (propagates across
it). Building the network is the expensive step, so if you need to
propagate many different seed sets across the *same* network, build it once
and pass it back in via `ig` instead of calling `netProp()` repeatedly:

```{r netprop-reuse, message=FALSE}
target_expr <- t(assay(demo_sciplex_se))
network <- wgcna.adj(target_expr, cor.type = "signed hybrid", diag_zero = TRUE, igraph = TRUE)

recon_netprop2 <- netProp(ig = network, seeds = demo_sciplex_sigs, sig = "rwr")
identical(recon_netprop, recon_netprop2)
```

See the [Network Propagation in Depth](2-network-propagation-details.html)
vignette for more on `wgcna.adj()`/`network_sig()`'s individual parameters.

## Method 3: Mean (DESeq2)

The mean method takes a different approach entirely: rather than
recontextualizing a signature you already have, it *derives* one directly
from target-context expression data, by running DESeq2 between perturbed and
control samples and keeping the top upregulated genes. It needs raw counts
and sample metadata identifying which samples are perturbed vs. control (and
by what), so it doesn't fit the SciPlex demo signatures used above -- here's
a minimal, fabricated example showing the shape of the inputs:

```{r mean-demo, message=FALSE}
set.seed(42)
n_genes <- 30

# g1/g2 are upregulated 4x in pertA samples; everything else is unperturbed noise.
ctrl_base <- rnbinom(n_genes, mu = 200, size = 10)
pert_base <- ctrl_base
pert_base[1:2] <- pert_base[1:2] * 4

sample_counts <- function(base, n) {
  sapply(seq_len(n), function(i) rnbinom(length(base), mu = pmax(base, 1), size = 10))
}

counts <- cbind(sample_counts(ctrl_base, 4), sample_counts(pert_base, 3))
rownames(counts) <- paste0("g", seq_len(n_genes))
colnames(counts) <- c(paste0("ctrl", 1:4), paste0("pertA", 1:3))

coldata <- S4Vectors::DataFrame(
  perturbation = c(rep("control", 4), rep("pertA", 3)),
  condition    = c(rep("Control", 4), rep("Perturbed", 3)),
  row.names    = colnames(counts)
)

se_mean <- SummarizedExperiment(assays = list(counts = counts), colData = coldata)

recon_mean <- recontextualize(
  method = "mean",
  se = se_mean,
  perturbation_col = "perturbation",
  condition_col = "condition",
  alpha = 1,
  limit = 2
)
recon_mean
```

`g1`/`g2` are the two genes consistently higher in `pertA` samples than in
controls, so they come out as the derived signature for `pertA`.

## One entry point: `recontextualize()`

Each method above (`projectCor()`, `netProp()`, the mean method) can also be
run through a single dispatcher, `recontextualize()`, by setting `method =
"projectCor"`, `"networkProp"`, or `"mean"`. This is convenient when method
choice is itself a parameter of a larger workflow (e.g. looping over
multiple methods in a benchmarking script) rather than something fixed at
call-site.

```{r recontextualize-equivalence}
identical(
  recontextualize(method = "projectCor", se = demo_sciplex_se, sigs = demo_sciplex_sigs, score = "gsva"),
  projectCor(demo_sciplex_se, demo_sciplex_sigs, score = "gsva")
)

identical(
  recontextualize(method = "networkProp", se = demo_sciplex_se, seeds = demo_sciplex_sigs, sig = "rwr"),
  netProp(demo_sciplex_se, seeds = demo_sciplex_sigs, sig = "rwr")
)

identical(
  recontextualize(
    method = "mean", se = se_mean,
    perturbation_col = "perturbation", condition_col = "condition",
    alpha = 1, limit = 2
  ),
  recon_mean
)
```

For `method = "networkProp"`, `recontextualize()` also accepts a pre-built
`ig` (just like `netProp()`) to skip network construction and reuse a
cached network:

```{r recontextualize-ig, message=FALSE}
identical(
  recontextualize(method = "networkProp", ig = network, seeds = demo_sciplex_sigs, sig = "rwr"),
  netProp(ig = network, seeds = demo_sciplex_sigs, sig = "rwr")
)
```

## Benchmarking against the ground truth

`sig_eval_table()` compares a recontextualized signature against both the
unmodified source signature and the true target-context signature, reporting:

- **`jacc`** -- Jaccard overlap between the recontextualized and true signature
- **`NES`** -- rank-based enrichment (via `fgsea`) of the recontextualized
  signature within the true signature's full ranked gene list
- **`displaced`/`kept`** -- how many source signature genes were replaced vs.
  retained during recontextualization

```{r benchmark}
eval_df <- sig_eval_table(
  source_sigs = demo_sciplex_sigs,
  pred_sigs   = recon_projectcor,
  true_sigs   = demo_sciplex_true_sigs,
  source      = "sciplex_k562",
  target      = "sciplex_a549"
)

head(eval_df[, c("gene", "jacc", "NES", "padj")])
```

### Is recontextualization actually helping?

The natural baseline to compare against is doing *nothing* -- using the
unmodified source signature directly as the "prediction." Computing the same
evaluation with `pred_sigs = demo_sciplex_sigs` gives that no-change
baseline:

```{r no-change-baseline}
no_change_df <- sig_eval_table(
  source_sigs = demo_sciplex_sigs,
  pred_sigs   = demo_sciplex_sigs,
  true_sigs   = demo_sciplex_true_sigs,
  source      = "sciplex_k562",
  target      = "sciplex_a549"
)

data.frame(
  method    = c("no_change", "projectCor"),
  mean_jacc = c(mean(no_change_df$jacc), mean(eval_df$jacc)),
  mean_NES  = c(mean(no_change_df$NES, na.rm = TRUE), mean(eval_df$NES, na.rm = TRUE))
)
```

On this demo, results are mixed: `projectCor()` improves average Jaccard
overlap with the true A549 signature, but average enrichment (NES) is
actually slightly *lower* than just reusing the K562 signature unmodified.
This is a realistic outcome, recontextualization does not
always outperform the no-change baseline for every method/metric/dataset
combination (see the paper for more details on benchmarking results).

`sigrecon::paired_eval_table()` formalizes
the no-change-vs-recontextualized comparison shown here, including paired
significance testing, for larger benchmarking workflows.

## Next Steps

- The [Network Propagation in Depth](2-network-propagation-details.html)
  vignette for `network_sig()`'s options
- The [BYOD](3-bring-your-own-signature.html) vignette for bringing your own
  perturbational data and signatures.
- The [Benchmarking Philosophy](4-benchmarking-and-splits.html) vignette for
  how we benchmark methods by taking into account target data availability.
