---
title: "Getting Started with sigRecon"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sigRecon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The signature recontextualization problem

Given a gene signature $X$ describing a genetic or chemical perturbation in one
biological context (a cell line, a tissue), what is the corresponding gene
signature of that *same* perturbation in a *different* context? Any method
that takes a ranked list of genes from a source context and outputs a ranked
list of genes for a target context is performing **signature
recontextualization**.

`sigrecon` provides two baseline recontextualization methods --
projection-based scoring (`projectCor()`) and network propagation
(`wgcna.adj()` + `network_sig()`) -- plus benchmarking tools
(`sig_eval_table()`) to evaluate how well a recontextualized signature
recovers the true signature in the target context.

This vignette walks through both methods end to end on a small, bundled
real-data example, requiring no downloads or external setup.

## Installation

```r
BiocManager::install("montilab/sigrecon", dependencies = TRUE)
```

```{r setup, message=FALSE}
library(sigrecon)
library(SummarizedExperiment)
```

## Demo Data

`sigrecon` bundles small, real-data recontextualization demos for each of the
datasets used in its benchmarking study (SciPlex, Tahoe, DrugMatrix,
Perturb-seq) -- see `?demo_datasets` for all four. This vignette uses the
SciPlex demo data:

- **`demo_sciplex_sigs`** -- SciPlex K562 signatures (top differentially
  expressed genes per drug), treated here as **source** signatures, as if
  defined in a different biological context than the one we're
  recontextualizing into.
- **`demo_sciplex_se`** -- SciPlex A549 expression data, representing
  the **target** context we want to recontextualize the source signatures
  into.
- **`demo_sciplex_true_sigs`** -- SciPlex A549 signatures for the same
  drugs, used as **ground truth** to check how well recontextualization
  worked.

```{r explore-demo}
dim(demo_sciplex_se)
length(demo_sciplex_sigs)
demo_sciplex_sigs[["Panobinostat (LBH589)"]][1:5]
```

Note `demo_sciplex_se` contains a *disjoint* set of drug profiles from those in
`demo_sciplex_sigs` and `demo_sciplex_true_sigs` -- so nothing
about the specific perturbations you're trying to recontextualize leaks into
the expression data used to predict new signatures.

## Method 1: Projection-based

`projectCor()` predicts a signature in a new context by scoring each
sample against the source signature (via GSVA or eigengene scoring) and then
ranking genes by their correlation with that per-sample score. The
highest-correlated genes become the recontextualized signature.

```{r projectcor}
recon_projectcor <- projectCor(demo_sciplex_se, demo_sciplex_sigs, score = "gsva")

# One recontextualized signature per source signature
length(recon_projectcor)
recon_projectcor[["Panobinostat (LBH589)"]][1:5]
```

## Method 2: Network Propagation

Network propagation works differently: `wgcna.adj()` first learns a gene
co-expression network from the target-context expression data, then
`network_sig()` propagates the source signature across that network (via
random walk with restart) to find the genes most "reachable" from the seed
genes -- these become the recontextualized signature.

```{r netprop, message=FALSE}
target_expr <- t(assay(demo_sciplex_se))
network <- wgcna.adj(target_expr, cor.type = "signed hybrid", diag_zero = TRUE, igraph = TRUE)

recon_netprop <- network_sig(network, seeds = demo_sciplex_sigs, sig = "rwr")
recon_netprop[["Panobinostat (LBH589)"]][1:5]
```

## Benchmarking against the ground truth

`sig_eval_table()` compares a recontextualized signature against both the
unmodified source signature and the true target-context signature, reporting:

- **`jacc`** -- Jaccard overlap between the recontextualized and true signature
- **`NES`** -- rank-based enrichment (via `fgsea`) of the recontextualized
  signature within the true signature's full ranked gene list
- **`displaced`/`kept`** -- how many source signature genes were replaced vs.
  retained during recontextualization

```{r benchmark}
eval_df <- sig_eval_table(
  source_sigs = demo_sciplex_sigs,
  pred_sigs   = recon_projectcor,
  true_sigs   = demo_sciplex_true_sigs,
  source      = "sciplex_k562",
  target      = "sciplex_a549"
)

head(eval_df[, c("gene", "jacc", "NES", "padj")])
```

### Is recontextualization actually helping?

The natural baseline to compare against is doing *nothing* -- using the
unmodified source signature directly as the "prediction." Computing the same
evaluation with `pred_sigs = demo_sciplex_sigs` gives that no-change
baseline:

```{r no-change-baseline}
no_change_df <- sig_eval_table(
  source_sigs = demo_sciplex_sigs,
  pred_sigs   = demo_sciplex_sigs,
  true_sigs   = demo_sciplex_true_sigs,
  source      = "sciplex_k562",
  target      = "sciplex_a549"
)

data.frame(
  method    = c("no_change", "projectCor"),
  mean_jacc = c(mean(no_change_df$jacc), mean(eval_df$jacc)),
  mean_NES  = c(mean(no_change_df$NES, na.rm = TRUE), mean(eval_df$NES, na.rm = TRUE))
)
```

On this demo, results are mixed: `projectCor()` improves average Jaccard
overlap with the true A549 signature, but average enrichment (NES) is
actually slightly *lower* than just reusing the K562 signature unmodified.
This is a realistic outcome, recontextualization does not
always outperform the no-change baseline for every method/metric/dataset
combination (see the paper for more details on benchmarking results).

`sigrecon::paired_eval_table()` formalizes
the no-change-vs-recontextualized comparison shown here, including paired
significance testing, for larger benchmarking workflows.

## Next Steps

- The [Network Propagation in Depth](2-network-propagation-details.html)
  vignette for `network_sig()`'s options
- The [BYOD](3-bring-your-own-signature.html) vignette for bringing your own
  perturbational data and signatures.
- The [Benchmarking Philosophy](4-benchmarking-and-splits.html) vignette for
  how we benchmark methods by taking into account target data availability.
