---
title: "Network Propagation in Depth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network Propagation in Depth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(sigrecon)
library(SummarizedExperiment)
```

The [Getting Started](1-getting-started.html) vignette uses `netProp()`, a
one-call wrapper around `wgcna.adj()` (builds the network) and
`network_sig()` (propagates across it) with their defaults. This vignette
goes deeper into those two pieces individually: how `wgcna.adj()` builds the
network `network_sig()` operates on, the two different algorithms behind
`sig = "corr"` vs. `sig = "rwr"`, and what each of `network_sig()`'s options
(`p`, `avg_p`, `bootstrap`, `n_bootstraps`, `limit`) actually controls.

Working with `wgcna.adj()`/`network_sig()` directly like this is worthwhile
whenever you want to reuse the *same* built network across many propagation
calls -- e.g. propagating many different seed sets, or sweeping over `p`
values, without repeating the expensive network-construction step each
time. `netProp()` supports this too, via its `ig` argument (see the end of
the Getting Started vignette) -- this vignette just shows what's underneath
that shortcut.

We'll use the bundled SciPlex demo data throughout.

```{r demo-data}
dim(demo_sciplex_se)
length(demo_sciplex_sigs)
```

## Building the Network

We found that using [WGCNA](https://link.springer.com/article/10.1186/1471-2105-9-559) works
well for signature recontextualization methods.

The `wgcna.adj()` helper function builds a weighted gene co-expression network from an 
expression matrix (samples as rows, genes as columns -- note this is **transposed**
relative to `SummarizedExperiment`'s gene-rows/sample-columns convention):

```{r build-network, message=FALSE}
target_expr <- t(assay(demo_sciplex_se))

network <- wgcna.adj(
  target_expr,
  cor.type  = "signed hybrid", # correlation sign handling -- see below
  diag_zero = TRUE,            # zero out self-loops
  igraph    = TRUE             # return an igraph object, not a raw matrix
)

network
```

A few of `wgcna.adj()`'s arguments worth knowing about:

- **`cor.type`**: `"unsigned"` treats positive and negative correlations as
  equally strong connections; `"signed"` only connects genes that move
  *together*; `"signed hybrid"` (used here) is a common middle ground --
  negative correlations are down-weighted rather than either kept at full
  strength or discarded entirely.
- **`cor.fn`**: `"bicor"` (biweight midcorrelation, the default) is more
  robust to outliers than Pearson (`"cor"`) -- relevant for pseudobulk
  expression data with a handful of replicates per condition.
- **`min.sft`/`beta`/`powers`**: WGCNA networks raise the correlation matrix
  to a soft-thresholding power to approximate a scale-free topology.
  `wgcna.adj()` automatically searches `powers` for the smallest one
  reaching `min.sft` fit; pass `beta` directly to skip that search if you
  already know the power you want.
- **`diag_zero`**: whether to zero the diagonal (self-correlation) before
  returning -- almost always `TRUE`, since a gene being perfectly correlated
  with itself is not informative for propagation.

## Propagation parameters

The examples below all propagate a single signature so the effect of each
parameter is easy to see in isolation:

```{r one-seed}
one_seed <- demo_sciplex_sigs[1]
names(one_seed)
```

### `p`: the restart probability

`p` controls how far the walk tends to travel from the seed genes before
teleporting back. Low `p` lets the walk wander further across the network
(more indirect, multi-hop genes get picked up); high `p` keeps it close to
the seeds (results converge toward the seeds' immediate neighborhood,
similar to `sig = "corr"`).

```{r restart-prob}
sig_p_low  <- network_sig(network, seeds = one_seed, sig = "rwr", p = 0.05, limit = 30)
sig_p_high <- network_sig(network, seeds = one_seed, sig = "rwr", p = 0.9, limit = 30)

jaccard(sig_p_low[[1]], sig_p_high[[1]])
```

### `avg_p`: ensembling over a range of restart values

Picking a single "right" `p` for a given network is not always obvious.
Setting `avg_p = TRUE` instead runs the random walk at `avg_p_length`
restart values evenly spaced across the `avg_p_vals` range and averages the
resulting stationary probabilities -- trading a single, possibly-arbitrary
choice of `p` for a smoother estimate less sensitive to that choice.

```{r avg-p}
sig_avg_p <- network_sig(
  network,
  seeds = one_seed,
  sig = "rwr",
  avg_p = TRUE,
  avg_p_vals = c(0.05, 0.5),
  avg_p_length = 5,
  limit = 30
)
length(sig_avg_p[[1]])
```

### `bootstrap`/`n_bootstraps`: significance-based selection instead of top-`limit`

By default (`bootstrap = FALSE`), `network_sig()` just keeps the top
`limit` genes by stationary probability -- an arbitrary cutoff. Setting
`bootstrap = TRUE` instead runs the *same* random walk from `n_bootstraps`
randomly-chosen gene sets (matched in size to the real seed set) to build an
empirical null distribution, then keeps only genes whose real stationary
probability exceeds the 99th percentile of that null -- a
significance-based cutoff rather than a fixed count. This means the
resulting signature can be smaller (or larger) than `limit`, and costs
roughly `n_bootstraps` extra random walks per seed set, so it's noticeably
slower.

```{r bootstrap}
sig_bootstrap <- network_sig(
  network,
  seeds = one_seed,
  sig = "rwr",
  bootstrap = TRUE,
  n_bootstraps = 30, # kept small here for vignette speed; real analyses may want more
  limit = 30 # unused when bootstrap = TRUE and p != 1, but harmless to leave set
)
length(sig_bootstrap[[1]])
```

## Next steps

- The [BYOD](3-bring-your-own-signature.html) vignette for bringing your own
  perturbational data and signatures.
- The [Benchmarking Philosophy](4-benchmarking-and-splits.html) vignette for
  how we benchmark methods by taking into account target data availability.
