---
title: "Bringing Your Own Signature and Expression Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bringing Your Own Signature and Expression Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(sigrecon)
library(SummarizedExperiment)
```

The [Getting Started](getting-started.html) vignette uses `sigrecon`'s
bundled demo data. This vignette instead walks through taking **your own**
differential expression (DE) results and expression data all the way through
`projectCor()`/`network_sig()` and `sig_eval_table()`.

## Extracting DEGs from a differential expression table

`sigrecon` represents a signature as a named list, one element per
perturbation, each a list with `up` (top significant upregulated genes) and
`up_full` (all genes, ranked). `sig_filter_fn()` builds this from a DE
results table.

### DESeq2 Table

`DESeq2::results()` output uses `log2FoldChange` and `padj` column names,
and typically has gene IDs as rownames rather than a column -- move them into
a column first. Here's a small simulated example with two perturbations:

```{r deseq2-example}
set.seed(1)
genes <- paste0("ENSG", sprintf("%06d", 1:40))

deseq2_table <- data.frame(
  ensembl_id     = rep(genes, times = 2),
  product        = rep(c("drugA", "drugB"), each = 40),
  log2FoldChange = c(rnorm(40, mean = 1), rnorm(40, mean = -1)),
  padj           = runif(80, 0, 0.2)
)

deseq2_sigs <- sig_filter_fn(
  deseq2_table,
  perts      = c("drugA", "drugB"),
  pert_col   = "product",
  log2fc_col = "log2FoldChange",
  pval_col   = "padj",
  geneid_col = "ensembl_id",
  limit      = 20
)

str(deseq2_sigs$drugA, max.level = 1)
```

### Seurat `FindMarkers()` Table 

Seurat's `FindMarkers()` output uses `avg_log2FC` and `p_val_adj` -- these
happen to be `sig_filter_fn()`'s defaults, so no column-name arguments are
needed if your table already has a perturbation-label column named
`"product"` and gene IDs as rownames-turned-column named `"ensembl_id"`
(otherwise, override `pert_col`/`geneid_col` the same way):

```{r seurat-example}
seurat_table <- data.frame(
  ensembl_id = rep(genes, times = 2),
  product    = rep(c("drugA", "drugB"), each = 40),
  avg_log2FC = c(rnorm(40, mean = 1), rnorm(40, mean = -1)),
  p_val_adj  = runif(80, 0, 0.2)
)

seurat_sigs <- sig_filter_fn(seurat_table, perts = c("drugA", "drugB"), limit = 20)
str(seurat_sigs$drugA, max.level = 1)
```

`sig_filter_fn()` can be used on any differential expression output (limma, DESeq2, Mast)
as long as you tell it which columns hold the perturbation label, log2 fold-change, p-value, and gene ID via `pert_col`/`log2fc_col`/`pval_col`/`geneid_col`.

## Next steps

- The [Getting Started](getting-started.html) vignette for a walkthrough
  using real bundled data, including a no-change baseline comparison
- `?demo_datasets` for real source/target signature pairs across four
  datasets, useful as a reference for what well-formed input looks like
