---
title: "Benchmarking: Controlling for target-data availability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking Design: Splits and Evaluation Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sigrecon)
```

The [Getting Started](1-getting-started.html) vignette runs `sig_eval_table()`
once and glances at a few of its output columns. This vignette covers the
evaluation design in full: what every output column means, the
Control/10th/90th split design used to benchmark methods at varying amounts of
target-context information, and how to compare against a no-change baseline
with `paired_eval_table()`.

## Understanding `sig_eval_table()` 

```{r eval-basic}
recon <- projectCor(demo_sciplex_se, demo_sciplex_sigs, score = "gsva")

eval_df <- sig_eval_table(
  source_sigs = demo_sciplex_sigs,
  pred_sigs   = recon,
  true_sigs   = demo_sciplex_true_sigs,
  source      = "sciplex_k562",
  target      = "sciplex_a549"
)

str(eval_df, max.level = 1)
```

Each row is one perturbation. Columns:

| Column | Meaning |
|---|---|
| `source`, `target` | Labels you passed in, identifying the source/target contexts |
| `gene` | The perturbation name (row identifier) |
| `displaced`, `kept` | How many source-signature genes were replaced vs. retained by recontextualization. `kept / (kept + displaced)` is the fraction of the source signature recontextualization left unchanged. |
| `jacc` | Jaccard overlap between the recontextualized signature and the true target-context signature's `up` genes |
| `ES`, `NES` | fgsea enrichment score / normalized enrichment score of the recontextualized signature against the true signature's full ranked gene list (`up_full`) -- how concentrated the recontextualized genes are near the top of the true ranking |
| `pval`, `padj` | Significance of that enrichment (raw and BH-adjusted) |
| `log2err`, `size` | fgsea's estimated log2 error on the p-value, and the number of recontextualized genes that overlapped the ranked list at all |
| `leadingEdge` | The specific genes driving the enrichment signal (a list-column -- one character vector per row) |

Higher `jacc` and `NES` mean the recontextualized signature more closely
resembles the true target-context signature -- i.e. better
recontextualization.

```{r leading-edge}
# leadingEdge is a list-column: one entry per row
eval_df$leadingEdge[[1]]
```

## The Control / 10th / 90th Benchmarking Regimes

Data-leakage is an important consideration for training and evaluating 
recontextualization methods. If methods are trained broadly on available 
perturbational data, as is often the case with foundation models, the 
true target-context perturbation signature may have already been seen at inference. 

One of the main contributions of `sigrecon` is to provide benchmarking methods 
for evaluating methods under three conditions that explicitly control how much 
target-context information is available:

- **`Control`**: None of the target perturbation profiles are available to the method.
- **`10th`**: 1/10th of the target perturbations are available. The method is then evaluated on
  the remaining 9/10th of held-out target perturbations.
- **`90th`**: 9/10th of the target perturbations are available. 
  The method is then evaluated on the remaining 1/10th of held-out target perturbations.

In the benchmarking scripts ([link](https://github.com/montilab/sigRecon_Benchmarking)), which perturbations fall in which split is decided once and saved to a `drug_splits.csv`/`pb_splits.csv` file. 

`sig_eval_table(splits = TRUE, split_file = ..., split_pb_col = ...,
split_type = c("10th", "90th"))` reads that file directly and filters which
perturbations get evaluated.

### An Example

`splits = TRUE` expects `pred_sigs` to be a *named list of splits*, each
element itself a signature list, with names matching the split file's
column names. Building a small one here from the demo signatures (reusing
the same predictions per split for simplicity -- in a real study, each
split's predictions come from a method run only on that split's reference
perturbations):

```{r build-splits}
set.seed(1)
drugs <- names(demo_sciplex_sigs)

split_tbl <- data.frame(
  drug     = drugs,
  split_1  = sample(c(TRUE, FALSE), length(drugs), replace = TRUE, prob = c(0.1, 0.9)),
  split_2  = sample(c(TRUE, FALSE), length(drugs), replace = TRUE, prob = c(0.1, 0.9))
)
table(split_tbl$split_1)

split_file <- tempfile(fileext = ".csv")
write.csv(split_tbl, split_file, row.names = FALSE)

pred_sigs_splits <- list(split_1 = recon, split_2 = recon)
```

```{r eval-10th}
eval_10th <- sig_eval_table(
  source_sigs  = demo_sciplex_sigs,
  pred_sigs    = pred_sigs_splits,
  true_sigs    = demo_sciplex_true_sigs,
  source       = "sciplex_k562",
  target       = "sciplex_a549",
  splits       = TRUE,
  split_file   = split_file,
  split_pb_col = "drug",
  split_type   = "10th"
)

# One extra "split" column identifying which split each row came from
table(eval_10th$split)
nrow(eval_10th)
```

`split_type = "10th"` evaluated only the perturbations where `split_1`/
`split_2` was `FALSE` (the withheld 9/10ths) -- fewer rows than the 23
perturbations in the full `demo_sciplex_sigs`, and split across the two
`split_1`/`split_2` conditions.

## Comparing against a no-change baseline

The natural baseline for "does recontextualization help?" is doing nothing --
using the unmodified source signature as the prediction. `sig_eval_table()`
computes this the same way as any other prediction, just passing the source
signatures as `pred_sigs`:

```{r no-change}
no_change_df <- sig_eval_table(
  source_sigs = demo_sciplex_sigs,
  pred_sigs   = demo_sciplex_sigs,
  true_sigs   = demo_sciplex_true_sigs,
  source      = "sciplex_k562",
  target      = "sciplex_a549"
)
```

`paired_eval_table()` formalizes the no-change-vs-recontextualized
comparison: it joins the two evaluation tables on `source`/`gene`, computes
`kept_alpha`/`NES_delta`/`Jacc_delta`, and prints a paired significance
summary (median performance, then a paired Wilcoxon test per source
combined across perturbations via Fisher's method). It takes the no-change
table as a **file path**, not a data frame in memory, so save it first:

```{r paired-eval}
no_change_path <- tempfile(fileext = ".rds")
saveRDS(no_change_df, no_change_path)

paired_df <- paired_eval_table(eval_df, no_change_path)
head(paired_df[, c("gene", "jacc_FALSE", "jacc_TRUE", "NES_FALSE", "NES_TRUE", "Jacc_delta", "NES_delta")])
```

`_FALSE` columns are the no-change baseline, `_TRUE` columns are the
recontextualized prediction (matching `sig_eval_table()`'s internal
`splits`/`filter_eval_inputs` naming convention, not a literal boolean
value) -- `Jacc_delta`/`NES_delta` being positive means recontextualization
improved on the baseline for that perturbation.

