---
title: "Bringing Your Own Signature and Expression Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bringing Your Own Signature and Expression Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sigrecon)
library(SummarizedExperiment)
```

The [Getting Started](getting-started.html) vignette uses `sigrecon`'s
bundled demo data. This vignette instead walks through taking **your own**
differential expression (DE) results and expression data all the way through
`projectCor()`/`network_sig()` and `sig_eval_table()`.

## 1. From a DE table to a signature

`sigrecon` represents a signature as a named list, one element per
perturbation, each a list with `up` (top significant upregulated genes) and
`up_full` (all genes, ranked). `sig_filter_fn()` builds this from a DE
results table.

### DESeq2-shaped input

`DESeq2::results()` output uses `log2FoldChange` and `padj` column names,
and typically has gene IDs as rownames rather than a column -- move them into
a column first. Here's a small simulated example with two perturbations:

```{r deseq2-example}
set.seed(1)
genes <- paste0("ENSG", sprintf("%06d", 1:40))

deseq2_table <- data.frame(
  ensembl_id     = rep(genes, times = 2),
  product        = rep(c("drugA", "drugB"), each = 40),
  log2FoldChange = c(rnorm(40, mean = 1), rnorm(40, mean = -1)),
  padj           = runif(80, 0, 0.2)
)

deseq2_sigs <- sig_filter_fn(
  deseq2_table,
  perts      = c("drugA", "drugB"),
  pert_col   = "product",
  log2fc_col = "log2FoldChange",
  pval_col   = "padj",
  geneid_col = "ensembl_id",
  limit      = 20
)

str(deseq2_sigs$drugA, max.level = 1)
```

### Seurat `FindMarkers()`-shaped input

Seurat's `FindMarkers()` output uses `avg_log2FC` and `p_val_adj` -- these
happen to be `sig_filter_fn()`'s defaults, so no column-name arguments are
needed if your table already has a perturbation-label column named
`"product"` and gene IDs as rownames-turned-column named `"ensembl_id"`
(otherwise, override `pert_col`/`geneid_col` the same way):

```{r seurat-example}
seurat_table <- data.frame(
  ensembl_id = rep(genes, times = 2),
  product    = rep(c("drugA", "drugB"), each = 40),
  avg_log2FC = c(rnorm(40, mean = 1), rnorm(40, mean = -1)),
  p_val_adj  = runif(80, 0, 0.2)
)

seurat_sigs <- sig_filter_fn(seurat_table, perts = c("drugA", "drugB"), limit = 20)
str(seurat_sigs$drugA, max.level = 1)
```

**The lesson**: `sig_filter_fn()` doesn't care which DE tool produced your
table -- DESeq2, `limma::topTable()`, Seurat's `FindMarkers()` (any test,
including MAST), or something else entirely -- as long as you tell it which
columns hold the perturbation label, log2 fold-change, p-value, and gene ID
via `pert_col`/`log2fc_col`/`pval_col`/`geneid_col`. Mixing DE tools across
datasets in the same analysis (e.g. DESeq2 for one cell line, MAST for
another) just means passing different arguments per call -- there's no need
to reconcile the tables into one schema first.

## 2. Gene ID namespace harmonization

A second, easy-to-miss issue: your DE table's gene IDs need to be in the
**same namespace** as whatever you're comparing against (a target signature,
or your expression data's rownames) -- both Ensembl IDs, or both HGNC gene
symbols, not a mix. `projectCor()`/`network_sig()` match genes by exact
string, so a namespace mismatch doesn't error -- it silently produces empty
or near-empty results.

If your DE table has gene symbols but you need Ensembl IDs (or vice versa),
join against a symbol/ID mapping table before calling `sig_filter_fn()`. A
common source for this mapping is an annotation package (e.g.
`org.Hs.eg.db`) or, if you built your expression data from a `Seurat`
object, its own feature metadata:

```{r id-mapping, eval = FALSE}
# Seurat objects often carry a symbol <-> Ensembl mapping in feature metadata
id_map <- seurat_obj@assays$RNA@meta.features |>
  tibble::rownames_to_column("ensembl_id") |>
  dplyr::select(ensembl_id, feature_name)

diff_table <- diff_table |>
  dplyr::left_join(id_map, by = c("gene_symbol" = "feature_name"))
```

Some symbols map to more than one Ensembl ID or vice versa; inspect
`id_map` for duplicates before joining, and decide how to resolve them (e.g.
keep the most-variable feature) rather than letting the join silently
fan out rows.

## 3. Recontextualizing your signature

Once you have a signature in the `up`/`up_full` format, the rest of the
workflow is the same as the [Getting Started](getting-started.html) vignette
-- just with your own expression data in place of the bundled demo. Building
a small simulated `SummarizedExperiment` over the same gene universe here to
keep this vignette self-contained:

```{r my-se}
n_samples <- 20
expr <- matrix(
  rnorm(length(genes) * n_samples, mean = 5), nrow = length(genes), ncol = n_samples,
  dimnames = list(genes, paste0("sample", 1:n_samples))
)

my_se <- SummarizedExperiment(
  assays  = list(logcounts = expr),
  colData = DataFrame(sample_id = colnames(expr))
)
```

`projectCor()`/`network_sig()`'s `sigs`/`seeds` arguments want just the
`up` gene vectors, not the full `up`/`up_full` list:

```{r recontextualize}
my_sigs_up <- lapply(deseq2_sigs, function(x) x$up)

recon_projectcor <- projectCor(my_se, my_sigs_up, score = "gsva")

my_mat <- t(assay(my_se))
my_network <- wgcna.adj(my_mat, cor.type = "signed hybrid", diag_zero = TRUE, igraph = TRUE)
recon_netprop <- network_sig(my_network, seeds = my_sigs_up, sig = "rwr")
```

## 4. Benchmarking

`sig_eval_table()`'s `true_sigs` argument expects the full `up`/`up_full`
shape (it needs the full ranked list to compute rank-based enrichment), so
pass the `sig_filter_fn()` output directly rather than the `up`-only list
used above:

```{r benchmark}
eval_df <- sig_eval_table(
  source_sigs = my_sigs_up,
  pred_sigs   = recon_projectcor,
  true_sigs   = seurat_sigs,
  source      = "my_source",
  target      = "my_target"
)

head(eval_df[, c("gene", "jacc", "NES", "padj")])
```

(This example uses `seurat_sigs` as a stand-in "ground truth" purely to
demonstrate the call shape -- in a real analysis, `true_sigs` would be an
independently-derived signature for the actual target context you're
recontextualizing into, not another simulated signature over the same random
data.)

## Next steps

- `?sig_filter_fn` for the full parameter reference
- The [Getting Started](getting-started.html) vignette for a walkthrough
  using real bundled data, including a no-change baseline comparison
- `?demo_datasets` for real source/target signature pairs across four
  datasets, useful as a reference for what well-formed input looks like
