---
title: "Network Propagation in Depth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network Propagation in Depth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sigrecon)
library(SummarizedExperiment)
```

The [Getting Started](getting-started.html) vignette uses `network_sig()` with
its defaults. This vignette goes deeper: how `wgcna.adj()` builds the
network `network_sig()` operates on, the two different algorithms behind
`sig = "corr"` vs. `sig = "rwr"`, and what each of `network_sig()`'s options
(`p`, `avg_p`, `bootstrap`, `n_bootstraps`, `limit`) actually controls.

We'll use the bundled SciPlex demo data throughout.

```{r demo-data}
dim(demo_sciplex_se)
length(demo_sciplex_sigs)
```

## Building the network with `wgcna.adj()`

`wgcna.adj()` builds a weighted gene co-expression network from an expression
matrix (samples as rows, genes as columns -- note this is **transposed**
relative to `SummarizedExperiment`'s gene-rows/sample-columns convention):

```{r build-network}
target_expr <- t(assay(demo_sciplex_se))

network <- wgcna.adj(
  target_expr,
  cor.type  = "signed hybrid", # correlation sign handling -- see below
  diag_zero = TRUE,            # zero out self-loops
  igraph    = TRUE             # return an igraph object, not a raw matrix
)

network
```

A few of `wgcna.adj()`'s arguments worth knowing about:

- **`cor.type`**: `"unsigned"` treats positive and negative correlations as
  equally strong connections; `"signed"` only connects genes that move
  *together*; `"signed hybrid"` (used here) is a common middle ground --
  negative correlations are down-weighted rather than either kept at full
  strength or discarded entirely.
- **`cor.fn`**: `"bicor"` (biweight midcorrelation, the default) is more
  robust to outliers than Pearson (`"cor"`) -- relevant for pseudobulk
  expression data with a handful of replicates per condition.
- **`min.sft`/`beta`/`powers`**: WGCNA networks raise the correlation matrix
  to a soft-thresholding power to approximate a scale-free topology.
  `wgcna.adj()` automatically searches `powers` for the smallest one
  reaching `min.sft` fit; pass `beta` directly to skip that search if you
  already know the power you want.
- **`diag_zero`**: whether to zero the diagonal (self-correlation) before
  returning -- almost always `TRUE`, since a gene being perfectly correlated
  with itself is not informative for propagation.

## `sig = "corr"` vs. `sig = "rwr"`: two different algorithms

`network_sig()` supports two mechanically distinct ways of finding a
seed signature's neighbors in the network:

- **`sig = "corr"`**: directly uses the network's edge weights. For each
  candidate gene, it averages that gene's correlation to every seed gene,
  then keeps the top `limit` genes by that average. This is a single
  lookup -- no iteration, no walking across the network -- so it's fast, but
  it only "sees" genes with a direct (first-degree) connection to the seeds.
- **`sig = "rwr"`** (random walk with restart): simulates a walker that
  starts at the seed genes and, at each step, either moves to a
  network-neighbor or teleports back to a seed gene (with probability `p`,
  see below). Run to convergence, this produces a **stationary probability**
  for every gene in the network -- how "reachable" it is from the seeds,
  accounting for indirect (multi-hop) paths. The top genes by stationary
  probability become the recontextualized signature.

```{r corr-vs-rwr}
one_seed <- demo_sciplex_sigs["Panobinostat (LBH589)"]

sig_corr <- network_sig(network, seeds = one_seed, sig = "corr", limit = 30)
sig_rwr  <- network_sig(network, seeds = one_seed, sig = "rwr", limit = 30)

# How much do the two methods agree on which genes are "close" to the seed?
length(intersect(sig_corr[[1]], sig_rwr[[1]]))
jaccard(sig_corr[[1]], sig_rwr[[1]])
```

The two methods substantially disagree here -- `sig = "rwr"` can surface
genes several hops away that `sig = "corr"` never considers, since `"corr"`
is restricted to genes directly connected to a seed gene by an edge. Neither
is "more correct" in general; which one recovers the true target-context
signature better is an empirical question this package's benchmarking
tools (`sig_eval_table()`) are built to answer per-dataset.

## `sig = "rwr"` options in depth

### `p`: the restart probability

`p` controls how far the walk tends to travel from the seed genes before
teleporting back. Low `p` lets the walk wander further across the network
(more indirect, multi-hop genes get picked up); high `p` keeps it close to
the seeds (results converge toward the seeds' immediate neighborhood,
similar to `sig = "corr"`).

```{r restart-prob}
sig_p_low  <- network_sig(network, seeds = one_seed, sig = "rwr", p = 0.05, limit = 30)
sig_p_high <- network_sig(network, seeds = one_seed, sig = "rwr", p = 0.9, limit = 30)

jaccard(sig_p_low[[1]], sig_p_high[[1]])
```

### `avg_p`: ensembling over a range of restart values

Picking a single "right" `p` for a given network is not always obvious.
Setting `avg_p = TRUE` instead runs the random walk at `avg_p_length`
restart values evenly spaced across the `avg_p_vals` range and averages the
resulting stationary probabilities -- trading a single, possibly-arbitrary
choice of `p` for a smoother estimate less sensitive to that choice.

```{r avg-p}
sig_avg_p <- network_sig(
  network,
  seeds = one_seed,
  sig = "rwr",
  avg_p = TRUE,
  avg_p_vals = c(0.05, 0.5),
  avg_p_length = 5,
  limit = 30
)
length(sig_avg_p[[1]])
```

### `bootstrap`/`n_bootstraps`: significance-based selection instead of top-`limit`

By default (`bootstrap = FALSE`), `network_sig()` just keeps the top
`limit` genes by stationary probability -- an arbitrary cutoff. Setting
`bootstrap = TRUE` instead runs the *same* random walk from `n_bootstraps`
randomly-chosen gene sets (matched in size to the real seed set) to build an
empirical null distribution, then keeps only genes whose real stationary
probability exceeds the 99th percentile of that null -- a
significance-based cutoff rather than a fixed count. This means the
resulting signature can be smaller (or larger) than `limit`, and costs
roughly `n_bootstraps` extra random walks per seed set, so it's noticeably
slower.

```{r bootstrap}
sig_bootstrap <- network_sig(
  network,
  seeds = one_seed,
  sig = "rwr",
  bootstrap = TRUE,
  n_bootstraps = 30, # kept small here for vignette speed; real analyses may want more
  limit = 30 # unused when bootstrap = TRUE and p != 1, but harmless to leave set
)
length(sig_bootstrap[[1]])
```

## Under the hood: `random_walk()` and seed matrices

`network_sig(sig = "rwr")` is itself built on the lower-level, exported
`random_walk()` function, which takes a **seed matrix** -- a
(genes x seed-sets) matrix marking which genes are seeds for which
signature -- rather than a named list of gene vectors. Building one by hand
shows exactly what `network_sig()` constructs internally:

```{r seed-matrix}
gene_names <- igraph::V(network)$name
seed_genes <- intersect(one_seed[[1]], gene_names)

seed_mat <- matrix(
  0, nrow = length(gene_names), ncol = 1,
  dimnames = list(gene_names, names(one_seed))
)
seed_mat[seed_genes, 1] <- 1

raw_probs <- random_walk(network, seed_mat, restart = 0.1)
dim(raw_probs)

# The top genes by stationary probability match extract_sig_mat()'s logic
# inside network_sig() -- this is the un-thresholded, raw output.
head(sort(raw_probs[, 1], decreasing = TRUE), 5)
```

`random_walk()` returns raw stationary probabilities for every gene, with no
top-`limit` cutoff or bootstrap-based thresholding applied -- that gene
selection step is what `network_sig()` adds on top.

## Next steps

- The [Getting Started](getting-started.html) vignette for the full
  recontextualization + benchmarking workflow
- `?network_sig`, `?random_walk`, `?wgcna.adj` for full parameter references
- `?demo_datasets` for the other three bundled demo triples
